Appendix

### 2025-10-17 manuscript_rmarkdown/manuscript_appendix.Rmd
Added data on residual errors for yeo-johnson transformation methods.

## Ranking Yeo-Johnson transformations (pt. II)

@@

# Add rank based on residual error for central normality.
data[outlier == FALSE, "method_rank" := rank(residuals), by = c("experiment_id", "outlier", "dataset")]
data[outlier == TRUE, "method_rank" := rank(residuals_central), by = c("experiment_id", "outlier", "dataset")]
data[outlier == FALSE, "norm_residuals" := residuals / n_samples]
data[outlier == TRUE, "norm_residuals" := residuals_central / n_samples]

@@

none                                  & & 0.246 & 0.139 & 0.186 & 0.125 & 0.187 & 0.125 \\
conventional                          & & 0.139 & 0.114 & 0.142 & 0.106 & 0.187 & 0.125 \\
conventional (z-score norm.)          & & 0.139 & 0.113 & 0.142 & 0.104 & 0.142 & 0.104 \\
conventional (robust scaling)         & & 0.139 & 0.114 & 0.141 & 0.104 & 0.141 & 0.104 \\
Raymaekers-Rousseeuw                  & & 0.211 & 0.121 & 0.183 & 1.212 & 0.187 & 0.125 \\
Raymaekers-Rousseeuw (z-score norm.)  & & 0.295 & 0.129 & 0.175 & 0.125 & 0.175 & 0.125 \\
Raymaekers-Rousseeuw (robust scaling) & & 0.209 & 0.124 & 0.172 & 0.116 & 0.172 & 0.116 \\
invariant                             & & 0.141 & 0.115 & 0.140 & 0.106 & 0.140 & 0.106 \\
robust invariant                      & & 0.141 & 0.100 & 0.140 & 0.095 & 0.140 & 0.090 \\

@@

none                                  & & 0.186 & 0.073 & 0.106 & 0.043 & 0.106 & 0.043 \\
conventional                          & & 0.014 & 0.060 & 0.060 & 0.046 & 0.106 & 0.043 \\
conventional (z-score norm.)          & & 0.021 & 0.061 & 0.070 & 0.040 & 0.070 & 0.040 \\
conventional (robust scaling)         & & 0.014 & 0.061 & 0.069 & 0.040 & 0.069 & 0.040 \\
Raymaekers-Rousseeuw                  & & 0.015 & 0.017 & 0.060 & 0.024 & 0.106 & 0.043 \\
Raymaekers-Rousseeuw (z-score norm.)  & & 0.021 & 0.016 & 0.070 & 0.024 & 0.070 & 0.024 \\
Raymaekers-Rousseeuw (robust scaling) & & 0.015 & 0.016 & 0.068 & 0.024 & 0.068 & 0.024 \\
invariant                             & & 0.013 & 0.067 & 0.059 & 0.044 & 0.059 & 0.044 \\
robust invariant                      & & 0.034 & 0.034 & 0.061 & 0.026 & 0.061 & 0.026 \\

>>>>

## Ranking Yeo-Johnson transformations: residual errors

In the main manuscript, transformation methods are compared by ranking within
each experiment, defined by the combination of sequence and absence/presence of
outliers. This allows for direct comparison of methods between different
experiments.

Table \ref{tab:comparison_methods_simulations_yeo_johnson_avg_residual} shows
the average normalised residual error for these transformation methods. This
metric is computed from the residual error and the residual error of the central
portion ($\kappa = 0.70$) for sequences without and with outliers, respectively.
In each experiment, the respective residual error was normalised by dividing with
the sequence length, and can be interpreted as the average residual error of a
single sample. This reduces the effect of sequence length on the residual error.

Table \ref{tab:comparison_methods_simulations_yeo_johnson_avg_residual_diff} shows
the average normalised difference in residual error between the best method for each
experiment and the individual methods. As above, normalisation was done by
division by the length of the sequence corresponding to each experiment.

Both tables show that the differences between best-performing methods on each
dataset are relatively small.

# Add rank based on residual error for central normality.
data[outlier == FALSE, "method_rank" := rank(residuals), by = c("experiment_id", "outlier", "dataset")]
data[outlier == TRUE, "method_rank" := rank(residuals_central), by = c("experiment_id", "outlier", "dataset")]
data[outlier == FALSE, "norm_residuals" := residuals / n_samples]
data[outlier == TRUE, "norm_residuals" := residuals_central / (0.8 * n_samples)]

data[outlier == FALSE, "min_residual" := min(residuals), by = c("experiment_id", "outlier", "dataset")]
data[outlier == TRUE, "min_residual" := min(residuals_central), by = c("experiment_id", "outlier", "dataset")]
data[outlier == FALSE, "residual_delta" := (residuals - min_residual) / n_samples]
data[outlier == TRUE, "residual_delta" := (residuals_central - min_residual) / (0.8 * n_samples)]

table_data <- data[
  ,
  list(
    "avg." = round(mean(method_rank), digits = 2L),
    "10%" = quantile(method_rank, 0.10),
    "50%" = quantile(method_rank, 0.50),
    "90%" = quantile(method_rank, 0.90),
    "avg. norm. res." = round(mean(norm_residuals), digits = 3),
    "avg. res. diff." = round(mean(residual_delta), digits = 3)
  ),
  by = c("transformation", "outlier", "dataset")
]

# Table with average rank --> this was already presented in the main manuscript.
data.table::dcast(transformation ~ dataset + outlier, data = table_data, value.var = "avg.")

# Table with difference with best-performing method in an experiment.
data.table::dcast(transformation ~ dataset + outlier, data = table_data, value.var = "avg. res. diff.")
```

\begin{table}
\begin{center}
\caption{
Comparison of average normalised residual error between Yeo-Johnson transformation methods based on either the residual error (without outliers) or the residual error of the central portion
(with outliers; $\kappa = 0.80$) over 3 datasets with 100000 sequences each. Residual errors were normalised by sequence length to reduce the effect of sequence length on the residual error.
The clean dataset consists of sequences derived through inverse Yeo-Johnson transformation
of data sampled from a standard normal distribution. The dirty dataset contains sequences sampled from asymmetric generalised normal distributions, centred at 0.
The shifted dataset also contains sequences sampled from asymmetric generalised normal distributions, but centred at 100, and scaled by 0.001.
Several transformation methods include normalisation before transformation, indicated by z-score normalisation (norm.) or robust scaling.
A lower normalised residual error is better. Discrepancies with the rank-based method may occur, as even with normalisation residual errors 
are not fully comparable between experiments, whereas the ranks achieved within each experiment are.
}
\label{tab:comparison_methods_simulations_yeo_johnson_avg_residual}
\begin{tabular}{l | l r r r r r r}

none                                  & & 0.202 & 0.112 & 0.128 & 0.080 & 0.129 & 0.080 \\
conventional                          & & 0.058 & 0.094 & 0.085 & 0.078 & 0.129 & 0.080 \\
conventional (z-score norm.)          & & 0.061 & 0.094 & 0.091 & 0.073 & 0.091 & 0.073 \\
conventional (robust scaling)         & & 0.058 & 0.094 & 0.090 & 0.073 & 0.090 & 0.073 \\
Raymaekers-Rousseeuw                  & & 0.071 & 0.055 & 0.092 & 0.279 & 0.129 & 0.080 \\
Raymaekers-Rousseeuw (z-score norm.)  & & 0.087 & 0.057 & 0.097 & 0.060 & 0.097 & 0.060 \\
Raymaekers-Rousseeuw (robust scaling) & & 0.071 & 0.056 & 0.095 & 0.059 & 0.095 & 0.059 \\
invariant                             & & 0.057 & 0.099 & 0.083 & 0.075 & 0.083 & 0.075 \\
robust invariant                      & & 0.069 & 0.066 & 0.084 & 0.058 & 0.084 & 0.058 \\

\begin{table}
\begin{center}
\caption{
Comparison of average difference in residual error between Yeo-Johnson
transformation methods and the best transformation method within each
experiment, normalised by the corresponding sequence length. This is based on
either the residual error (without outliers) or the residual error of the
central portion (with outliers; $\kappa = 0.80$) over 3 datasets with 100000
sequences each. Residual errors were normalised by sequence length to reduce the
effect of sequence length on the residual error. The clean dataset consists of
sequences derived through inverse Yeo-Johnson transformation of data sampled
from a standard normal distribution. The dirty dataset contains sequences
sampled from asymmetric generalised normal distributions, centred at 0. The
shifted dataset also contains sequences sampled from asymmetric generalised
normal distributions, but centred at 100, and scaled by 0.001. Several
transformation methods include normalisation before transformation, indicated by
z-score normalisation (norm.) or robust scaling. A lower difference is better,
with a minimum value of 0.0. Discrepancies with the rank-based method may occur,
as normalised differences in residual error are not fully comparable between
experiments, whereas the ranks achieved within each experiment are.
}
\label{tab:comparison_methods_simulations_yeo_johnson_avg_residual_diff}
\begin{tabular}{l | l r r r r r r}

none                                  & & 0.150 & 0.068 & 0.050 & 0.033 & 0.049 & 0.033 \\
conventional                          & & 0.005 & 0.049 & 0.006 & 0.032 & 0.049 & 0.033 \\
conventional (z-score norm.)          & & 0.008 & 0.049 & 0.013 & 0.026 & 0.011 & 0.025 \\
conventional (robust scaling)         & & 0.005 & 0.050 & 0.012 & 0.026 & 0.010 & 0.026 \\
Raymaekers-Rousseeuw                  & & 0.018 & 0.011 & 0.014 & 0.233 & 0.049 & 0.033 \\
Raymaekers-Rousseeuw (z-score norm.)  & & 0.034 & 0.012 & 0.018 & 0.014 & 0.017 & 0.013 \\
Raymaekers-Rousseeuw (robust scaling) & & 0.018 & 0.011 & 0.017 & 0.012 & 0.015 & 0.011 \\
invariant                             & & 0.004 & 0.054 & 0.005 & 0.029 & 0.003 & 0.028 \\
robust invariant                      & & 0.016 & 0.021 & 0.006 & 0.011 & 0.004 & 0.010 \\

## Ranking Yeo-Johnson transformations: effect of sequence length

@@

# Add rank based on residual error for central normality.
data[outlier == FALSE, "method_rank" := rank(residuals), by = c("experiment_id", "outlier", "dataset")]
data[outlier == TRUE, "method_rank" := rank(residuals_central), by = c("experiment_id", "outlier", "dataset")]
data[outlier == FALSE, "norm_residuals" := residuals / n_samples]
data[outlier == TRUE, "norm_residuals" := residuals_central / (0.8 * n_samples)]

@@

none                                  & & 0.246 & 0.173 & 0.186 & 0.156 & 0.187 & 0.157 \\
conventional                          & & 0.139 & 0.142 & 0.142 & 0.132 & 0.187 & 0.157 \\
conventional (z-score norm.)          & & 0.139 & 0.142 & 0.142 & 0.130 & 0.142 & 0.130 \\
conventional (robust scaling)         & & 0.139 & 0.142 & 0.141 & 0.130 & 0.141 & 0.130 \\
Raymaekers-Rousseeuw                  & & 0.211 & 0.151 & 0.183 & 1.515 & 0.187 & 0.157 \\
Raymaekers-Rousseeuw (z-score norm.)  & & 0.295 & 0.161 & 0.175 & 0.156 & 0.175 & 0.156 \\
Raymaekers-Rousseeuw (robust scaling) & & 0.209 & 0.155 & 0.172 & 0.144 & 0.172 & 0.144 \\
invariant                             & & 0.141 & 0.144 & 0.140 & 0.132 & 0.140 & 0.132 \\
robust invariant                      & & 0.141 & 0.125 & 0.140 & 0.119 & 0.140 & 0.119 \\

@@

none                                  & & 0.186 & 0.091 & 0.106 & 0.053 & 0.106 & 0.053 \\
conventional                          & & 0.014 & 0.076 & 0.060 & 0.057 & 0.106 & 0.053 \\
conventional (z-score norm.)          & & 0.021 & 0.076 & 0.070 & 0.050 & 0.070 & 0.050 \\
conventional (robust scaling)         & & 0.014 & 0.076 & 0.069 & 0.050 & 0.069 & 0.050 \\
Raymaekers-Rousseeuw                  & & 0.015 & 0.021 & 0.060 & 0.030 & 0.106 & 0.053 \\
Raymaekers-Rousseeuw (z-score norm.)  & & 0.021 & 0.020 & 0.070 & 0.030 & 0.070 & 0.030 \\
Raymaekers-Rousseeuw (robust scaling) & & 0.015 & 0.020 & 0.068 & 0.030 & 0.068 & 0.030 \\
invariant                             & & 0.013 & 0.084 & 0.059 & 0.055 & 0.059 & 0.055 \\
robust invariant                      & & 0.034 & 0.042 & 0.061 & 0.032 & 0.061 & 0.032 \\
